---
title: "Sperm Motility Analysis of *Atelopus* sp. under Cryoprotectant Formulations"
author: "Andrés Benjamin Garcés Cifuentes"
date: "`r Sys.Date()`"
output: html_document:
  toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    fig_caption: true
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```
```{r libraries}
library(tidyverse)
library(readxl)
library(pwr)
library(ggdist)
library(knitr)
library(kableExtra)
```
## Background

This report presents a statistical analysis of sperm motility data collected
from *Atelopus* sp. specimens at Fundación Jambatu as part of an ongoing
amphibian cryopreservation research programme. The dataset used here is a
simulated representation of pilot experimental findings, produced to protect
confidential project information while preserving the statistical
characteristics of the original data.

Two cryoprotectant formulations (CPA1 and CPA2) were evaluated across four
time lapses (5, 10, 15, and 20 minutes post-activation). The primary outcome
measure is the percentage of motile sperm at each time point.

## Data
```{r load-data}
xlr <- read_xlsx("prd.xlsx", range = "A1:C40", col_names = FALSE)
colnames(xlr) <- c("Treatment", "Time_Lapse", "Motility")

xlr$Time_Lapse <- factor(xlr$Time_Lapse,
                         levels = c(5, 10, 15, 20))

cpa1r <- xlr %>% filter(Treatment == "CPA1")
cpa2r <- xlr %>% filter(Treatment == "CPA2")

cpa1_summary <- cpa1r %>%
  group_by(Time_Lapse) %>%
  summarize(Mean = mean(Motility, na.rm = TRUE),
            SD   = sd(Motility,   na.rm = TRUE),
            N    = n())

cpa2_summary <- cpa2r %>%
  group_by(Time_Lapse) %>%
  summarize(Mean = mean(Motility, na.rm = TRUE),
            SD   = sd(Motility,   na.rm = TRUE),
            N    = n())
```

### Descriptive statistics
```{r summary-tables}
kable(cpa1_summary,
      digits  = 2,
      caption = "Table 1. Mean motility (%) per time lapse — CPA1") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

kable(cpa2_summary,
      digits  = 2,
      caption = "Table 2. Mean motility (%) per time lapse — CPA2") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```
## Visualisation

Raincloud plots combine a half-violin (distribution shape), a box plot
(median and IQR), and the raw jittered data points. This format is
recommended for small samples because it communicates distribution shape
and individual variation that bar plots hide (Allen et al., 2019;
Weissgerber et al., 2015).
```{r raincloud, fig.cap="Figure 1. Sperm motility across time lapses for CPA1 and CPA2. Half-violins show the distribution; box plots show the median and IQR; points show individual observations."}
pal <- c("CPA1" = "#2E6F8E", "CPA2" = "#1D6A52")
```

## Statistical analysis

### Two-way ANOVA
A two-way ANOVA was fitted with Treatment and Time_Lapse as fixed factors
and their interaction term, using Motility as the continuous dependent
variable. The null hypotheses are: (1) Treatment has no effect on mean
motility; (2) Time_Lapse has no effect on mean motility; (3) the effect
of time does not differ between treatments.

```{r two-way-anova}
anova_two_way <- aov(Motility ~ Treatment * Time_Lapse, data = xlr)

kable(broom::tidy(anova_two_way),
      digits  = 4,
      caption = "Table 3. Two-way ANOVA results.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

Treatment alone did not produce a significant difference in mean motility
between CPA1 and CPA2. Time_Lapse was significant, indicating motility
changes over time regardless of cryoprotectant. The non-significant
interaction term suggests both formulations follow a similar
time-dependent trajectory.

ggplot(xlr, aes(x = Time_Lapse, y = Motility,
                fill = Treatment, colour = Treatment)) +
  stat_halfeye(adjust = 0.5, width = 0.4, justification = -0.3,
               .width = 0, point_colour = NA, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               alpha = 0.5, colour = "grey30") +
  geom_jitter(width = 0.05, alpha = 0.6, size = 1.8) +
  scale_fill_colour_manual(values = pal) +
  facet_wrap(~ Treatment, ncol = 2) +
  scale_y_continuous(limits = c(40, 110),
                     breaks = seq(40, 100, 20)) +
  labs(x = "Time lapse (minutes)", y = "Motility (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        strip.text       = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

### One-way ANOVA per treatment

To determine whether motility differs significantly across time lapses
within each formulation independently, one-way ANOVAs were fitted
separately for CPA1 and CPA2.
```{r one-way-anova}
anova_cpa1 <- aov(Motility ~ factor(Time_Lapse), data = cpa1r)
anova_cpa2 <- aov(Motility ~ factor(Time_Lapse), data = cpa2r)

kable(broom::tidy(anova_cpa1),
      digits  = 4,
      caption = "Table 4. One-way ANOVA — CPA1.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

kable(broom::tidy(anova_cpa2),
      digits  = 4,
      caption = "Table 5. One-way ANOVA — CPA2.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

CPA1 showed no significant time effect (p > 0.05). CPA2 showed a
significant time effect (p < 0.05), warranting post-hoc testing.

### Tukey HSD post-hoc test — CPA2
```{r tukey}
model_cpa2 <- aov(Motility ~ factor(Time_Lapse), data = cpa2r)

tukey_res <- TukeyHSD(model_cpa2)

kable(broom::tidy(tukey_res),
      digits  = 4,
      caption = "Table 6. Tukey HSD pairwise comparisons — CPA2.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(which(broom::tidy(tukey_res)$adj.p.value < 0.05),
           bold = TRUE, color = "white", background = "#1D6A52")
```

Significant declines in motility were observed between Time 5 and Time 15,
and between Time 5 and Time 20 (p adj < 0.05, highlighted above).
All other pairwise comparisons were non-significant, suggesting motility
stabilises after the initial drop.