---
title: "Sperm Motility Analysis of *Atelopus* sp. under Cryoprotectant Formulations"
author: "Andrés Benjamin Garcés Cifuentes"
date: "`r Sys.Date()`"
output: html_document:
  toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    fig_caption: true
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```
```{r libraries}
library(tidyverse)
library(readxl)
library(pwr)
library(ggdist)
library(knitr)
library(kableExtra)
```
## Background

This report presents a statistical analysis of sperm motility data collected
from *Atelopus* sp. specimens at Fundación Jambatu as part of an ongoing
amphibian cryopreservation research programme. The dataset used here is a
simulated representation of pilot experimental findings, produced to protect
confidential project information while preserving the statistical
characteristics of the original data.

Two cryoprotectant formulations (CPA1 and CPA2) were evaluated across four
time lapses (5, 10, 15, and 20 minutes post-activation). The primary outcome
measure is the percentage of motile sperm at each time point.

## Data
```{r load-data}
xlr <- read_xlsx("prd.xlsx", range = "A1:C40", col_names = FALSE)
colnames(xlr) <- c("Treatment", "Time_Lapse", "Motility")

xlr$Time_Lapse <- factor(xlr$Time_Lapse,
                         levels = c(5, 10, 15, 20))

cpa1r <- xlr %>% filter(Treatment == "CPA1")
cpa2r <- xlr %>% filter(Treatment == "CPA2")

cpa1_summary <- cpa1r %>%
  group_by(Time_Lapse) %>%
  summarize(Mean = mean(Motility, na.rm = TRUE),
            SD   = sd(Motility,   na.rm = TRUE),
            N    = n())

cpa2_summary <- cpa2r %>%
  group_by(Time_Lapse) %>%
  summarize(Mean = mean(Motility, na.rm = TRUE),
            SD   = sd(Motility,   na.rm = TRUE),
            N    = n())
```

### Descriptive statistics
```{r summary-tables}
kable(cpa1_summary,
      digits  = 2,
      caption = "Table 1. Mean motility (%) per time lapse — CPA1") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

kable(cpa2_summary,
      digits  = 2,
      caption = "Table 2. Mean motility (%) per time lapse — CPA2") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```